---
title: "visualization"
subtitle: "part 1"
output: github_document
---
install.packages("ggridges")

```{r}
library (tidyverse)
library (ggridges)
library (rnoaa)
```

## load the weather data
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match (
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything()) # put name and id in the front, everthing comes follow

```

## scatterplots!!

create my first scatterplot ever.

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point() #generate the points
```

New approach, same plot.
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```

save abd edur a plot object.
```{r}
weather_plot = 
  weather_df |>
  ggplot(aes(x = tmin, y = tmax))

weather_plot + geom_point ()
```

## Advance scatterplot ...

start with the same one and make it fancy!

```{r}
warther_df |>
  ggplot (aes(x = tmin, y = tmax, color = name)) + #points become colorful, depend on different categories, points are labeled into different colors
  geom_point() + #show points
  geom_smooth (se = FALSE) #standard error = false, show 3 smooth curves between points, indicates se for 3 variables #show lines
```

what about the `aes` placement?
```{r}
warther_df |>
  ggplot (aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name)) + # color only define inside of geom_point instead of the whole things
  geom_smooth () #where is the different. only one line for the whole graph
```

Let's facet some things!!

```{r}
warther_df |>
  ggplot (aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.2, size = 0.3) + #define transparency, make points transparent, so that you can see points and line, small number = more transparent
  geom_smooth (se = FALSE, size = 2) + 
  facet_grid(. ~ name) #generate 3 planes plots for 3 variables, vertical

#if put facet_grid (name ~ .) plots became horizontal 

warther_df |>
  ggplot (aes(x = tmin, y = tmax, alpha = tmin, color = name)) + #alpha = tmin, make points, transparent on the bottom, visible on the top, 
  geom_point() + 
  geom_smooth (se = FALSE) + 
  facet_grid(. ~ name) 

```

Let's combine some elements and try a new plot.
```{r}
weather_df |>
  ggplot (aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = 0.5) + #aes(size = prcp) based on the intensity, points big or small
  geom_smooth (se = FALSE)

```

## some small notes

how mant geoms have to exist?
- you can have whatever geom you want.
```{r}
waether_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_smooth (se = FALSE)
```


You can use a neat geom!
```{r}
weather_df |>
  ggplot(aes (x = tmin, y = tmax)) + 
  geom_hex() #points become hexagon

#geom_bin2d() points become square
#geom_density2d () points become contour plot
```

## Univariate plots
histograms are really great.
```{r}
weather_df |>
  ggplot (aes(x=tmin)) +
  geom_histogram()
```

Can we add color ..
```{r}
weather_df |>
  ggplot (aes(x=tmin, fill = name)) +
  geom_histogram(position = "dodge") # bars list side by side
```

```{r}
weather_df |>
  ggplot (aes(x=tmin, fill = name)) +
  geom_histogram() +
  facet_grid (.~ name) # generate 3 plots respectively 
```

Let's try new geometry!
```{r}
waether_df |>
  ggplot (ase (x = tmin, fil= name)) +
  geom_density (alpha = 0.3, adjust = .5)
```

Box plots

```{r}
weather_df |>
  ggplot (aes(x = name, y = tmin)) +
  geom_boxplot()
```


Trendy plots

```{r}
weather_df |>
  ggplot(aes(x = name, y = tmin, fill = name)) |>
  geom_violin (alpha = .5) +
  stat_summary (fun = "median")
```


Ridge plots - - the most popular plot of 2017
```{r}
weather_df |>
  ggplot (aes(x = tmin, y = name)) +
  geom_density_ridges ()
```

## save and embed

Let's save a scatterplot
```{r}
weather_plot |>
  ggplot (aes (x = tmin, y = tmax, color = name)) +
  gemo_point (alpha = .5)

ggsave ("./results/weather_plot.pdf", weather_plot, width = 8, height = 5)
```

What about embedding 
```{r, fig.width= 12, fig.height=6}
weather_plot
```





